---
title: "equity_analysis"
author: "Richard Barad"
date: "2024-06-10"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(sf)
library(tidycensus)
library(readxl)
library(tigris)
library(httr)
library(kableExtra)
library(viridis)
library(scales)

```

## Read Project Data

Read CWSRF Project data

```{r read_data}

data_raw <- read_excel('Data/De_river_states.xlsx') %>%
  mutate(total = Nature_Based + Gray,
         stormwater = Green_Infrastructure + Gray_Infrastructure,
         wastewater = Advanced_Treatment + CSO_Correction + Infiltration_Inflow_Correction + New_Collector_Sewers + New_Interceptors + Secondary_Treatment + Sewer_System_Rehabilitation + Planning_and_Assessments,
         energy_conservation = Energy_Efficiency + Renewable_Energy,
         water_conservation = Water_Efficiency,
         equity_analysis_funding = stormwater + wastewater + energy_conservation + water_conservation) 

data <- data_raw %>%
  st_as_sf(., coords = c("Longitude","Latitude"),crs='EPSG:4326') %>%
  st_transform('EPSG:2272')

```

## Get Urban Areas and map

Get DE Basin boundaries, urban areas using tigris package and intersect urban areas with states to get just urban areas in states that are part of study area.

```{r urban_areas, echo=FALSE}

de_basin <- st_read('Data/DE_Basin.shp') %>% st_transform('EPSG:2272')

#Get Urban Areas and intersect with States

urban_areas <- tigris::urban_areas(year=2020,criteria=2020)

states <- tigris::states() %>%
  dplyr::filter(NAME %in% c('Pennsylvania','New Jersey','Delaware'))

states_2272 <- states %>% st_transform('EPSG:2272')

urban_areas <- st_intersection(states,urban_areas) %>% st_transform('EPSG:2272')

#Map quick map

bbox <- st_bbox(de_basin)

xmin <- bbox[1]
ymin <- bbox[2]
xmax <- bbox[3]
ymax <- bbox[4]

ggplot()+
  geom_sf(data=urban_areas,fill='orange',color='transparent')+
  geom_sf(data=states_2272,color='gray40',fill='transparent')+
  geom_sf(data=de_basin,color='lightblue',fill='transparent',linewidth=1)+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()

ggsave('outputs/Equity_Analysis/urban_areas2.png',height=3.8,width=2)

```

# Get County Boundaries and Determine Counties in Study Area

Get county boundaries for states of interest, and determine counties that have an area overlap with the study area that is greater than 40%. These counties will be considered part of the study area. 

``` {r counties}

counties <- rbind(counties(state='PA'),counties(state='DE'),counties(state='NJ')) %>% st_transform('EPSG:2272')

# Include counties that have a more 40% overlap with the Basin

basin_counties <- counties %>%
  mutate(area = st_area(.)) %>%
  st_intersection(.,de_basin) %>%
  mutate(basin_area = st_area(.),
         percent_basin = as.numeric(basin_area) / as.numeric(area)) %>%
  filter(percent_basin > 0.4) %>%
  st_drop_geometry() %>%
  select('GEOID','STATEFP','COUNTYFP','NAME') %>%
  inner_join(counties %>% select('geometry','GEOID'),.,by='GEOID')

ggplot()+
  geom_sf(data=states,color='gray50',size=2,fill='transparent')+
  geom_sf(data=basin_counties,fill='#ff9b9b',color='gray60')+
  geom_sf(data=states,fill='transparent')+
  theme_void()

ggsave('outputs/Equity_Analysis/study_area.png')

```


# Get Population Data

Get population data 

```{r get_population}

# Get Population Data for counties in study area

pop_PA <- get_decennial('block',sumfile='dhc',year=2020,state='PA',variable='P1_001N',geometry=TRUE,county=c('Wayne','Philadelphia','Berks','Schuylkill','Chester','Monroe','Carbon','Montgomery','Delaware','Lehigh','Northampton','Bucks','Pike'))
pop_DE <- get_decennial('block',sumfile='dhc',year=2020,state='DE',variable='P1_001N',geometry=TRUE,county=c('New Castle','Kent'))
pop_NJ <- get_decennial('block',sumfile='dhc',year=2020,state='NJ',variable='P1_001N',geometry=TRUE,county=c('Sussex','Camden','Cape May','Burlington','Gloucester','Cumberland','Warren','Salem','Mercer'))

pop <- rbind(pop_PA,pop_DE,pop_NJ) %>% st_transform('EPSG:2272') %>%
  separate(.,col=NAME,into=c('Block','Block Group','Tract','County','State'),sep=', ')

pop$NAME <- pop$County %>% stringr::str_remove(" County")

basin_blocks <- pop %>% 
  st_centroid()

#Determine which census blocks are urban and which are rural - use centroid of census block group

urban_blocks <- basin_blocks[st_intersects(basin_blocks,urban_areas) %>% lengths > 0, ] %>% mutate(urban = 'Urban')
rural_blocks <- basin_blocks[st_intersects(basin_blocks,urban_areas) %>% lengths == 0, ] %>% mutate(urban = 'Rural')

#Merge and Join back to polygon geometry

pop_data_basin <- rbind(urban_blocks,rural_blocks) %>%
  st_drop_geometry() %>%
  left_join(.,pop %>% select('GEOID','geometry'),by='GEOID') %>%
  st_as_sf()
```

# Get CJEST DAta

``` {r CJEST}

#CJEST Shapefile is available here: https://screeningtool.geoplatform.gov/en/downloads

CEJST_burdened <- st_read('C:/Users/richa/OneDrive/Documents/Professional/Water Center Work/CEJST/usa.shp') %>%
  filter(SF %in% c('New Jersey','Delaware','Pennsylvania')) %>%
  st_transform('EPSG:2272') %>%
  filter(SN_C == 1)

pop_centroid <- pop_data_basin %>% st_centroid()

#Determine which census blocks are overburdened - use centroid of census block group

burdened_blocks <- pop_centroid[st_intersects(pop_centroid,CEJST_burdened) %>% lengths > 0, ] %>% mutate(overburdened = 'Yes')
nonburdned_blocks <- pop_centroid[st_intersects(pop_centroid,CEJST_burdened) %>% lengths == 0, ] %>% mutate(overburdened = 'No')

#Merge and Join back to polygon geometry

pop_data_basin <- rbind(burdened_blocks,nonburdned_blocks) %>%
  st_drop_geometry() %>%
  left_join(.,pop_data_basin %>% select('GEOID','geometry'),by='GEOID') %>%
  st_as_sf() %>%
  mutate(area_ft = as.numeric(st_area(.)))

```

# Map Rural / Urban Areas

```{r}
bbox <- st_bbox(pop_data_basin)

xmin <- bbox[1]
ymin <- bbox[2]
xmax <- bbox[3]
ymax <- bbox[4]

ggplot()+
  geom_sf(data=states_2272,fill='gray90')+
  geom_sf(data=pop_data_basin,aes(fill=as.factor(urban)),color='transparent')+
  scale_fill_manual(values=c('lightgreen','pink'),labels=c('Rural','Urban'),name='')+
  geom_sf(data=basin_counties,color='gray60',fill='transparent',size=0.5)+
  geom_sf(data=states_2272,color='gray40',fill='transparent')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2.5)+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()

ggsave('outputs/Equity_Analysis/urban_areas.png',width=3,height=5)

```

``` {r join_project_funding}

#Determine Total CWSRF Funding By County

funding_county <- st_join(data,basin_counties,left=FALSE) %>%
  filter(score > 95) %>% #Drop projects that do not have a specific geographic location
  group_by(NAME) %>% summarise(funding = sum(equity_analysis_funding),subsidy=sum(Additional_Subsidy_Amount)) %>%
  st_drop_geometry()

Data_County1 <- left_join(basin_counties,funding_county,by='NAME') %>%
  st_as_sf() %>%
  select(NAME,GEOID,funding,subsidy) %>%
  mutate(funding = ifelse(is.na(funding),0,funding),
         subsidy = ifelse(is.na(subsidy),0,subsidy))

```

``` {r join_urban_pop}

#Determine Rural / Urban Population by County and Rural / Urban Area by County

pop_county <- pop_data_basin %>%
  st_drop_geometry() %>%
  group_by(NAME,urban,State) %>% summarize(pop = sum(value),area=sum(area_ft)) %>%
  pivot_wider(id_cols=c(NAME,State),names_from=urban,values_from=c(pop,area)) %>%
  mutate(pop_Rural = ifelse(is.na(pop_Rural),0,pop_Rural),
         area_Rural = ifelse(is.na(area_Rural),0,area_Rural),
         percent_urban_pop = round(pop_Urban / (pop_Urban + pop_Rural),4),
         percent_urban_area = round(area_Urban / (area_Urban + area_Rural),4)) %>%
  select(NAME,State,pop_Urban,pop_Rural,percent_urban_pop,percent_urban_area)

#Merge with funding by county data and calculate per capita information
Data_County2 <- left_join(Data_County1,pop_county,by='NAME') %>%
  mutate(funding_per_urban_capita = funding / pop_Urban,
         subsidy_per_capita = subsidy / pop_Urban)

```

```{r join_overburdned}

#Calculate percent of urban population that is also overburdened according to CJEST

overburdened <- pop_data_basin %>% filter(urban == 'Urban') %>%
  st_drop_geometry() %>%
  group_by(NAME,overburdened) %>% summarize(pop = sum(value)) %>%
  pivot_wider(id_cols=NAME,names_from=overburdened,values_from=pop) %>%
  mutate(Yes = ifelse(is.na(Yes),0,Yes),
          percent_urban_overbudened = Yes / (Yes + No)) %>%
  select(NAME,percent_urban_overbudened)


Data_County3 <- left_join(Data_County2,overburdened,by='NAME')

Data_County3_centroid <- Data_County3 %>% st_centroid()

```


``` {r make_maps}

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=funding),color='gray80')+
  scale_fill_distiller(palette='Greens',direction=1,labels = dollar, name='Financing ($)')+
  geom_sf(data=states_2272,fill='transparent',color='gray30')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='black')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()

ggsave('outputs/Equity_Analysis/financing.png')

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=funding_per_urban_capita),color='gray80')+
  scale_fill_distiller(palette='Greens',direction=1,name='Financing ($) Per \n Urban Capita')+
  geom_sf(data=states_2272,fill='transparent',color='gray30')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='black')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()+
  theme(legend.position="bottom",legend.text=element_text(size=6))

ggsave('outputs/Equity_Analysis/financing_per_capita.png')

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=pop_Urban / 1000),color='gray80')+
  scale_fill_viridis(option='rocket',direction=-1,name='Urban Population \n (thousands)')+
  geom_sf(data=states_2272,fill='transparent',color='gray30')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='black')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()+
  theme(legend.position="bottom",legend.text=element_text(size=9))

ggsave('outputs/Equity_Analysis/urban_pop.png')

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=percent_urban_pop * 100),color='gray80')+
  scale_fill_viridis(option='rocket',direction=-1,name='Percent Urban')+
  geom_sf(data=states_2272 %>% st_transform('EPSG:3857'),fill='transparent',color='gray30')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='white')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()+
  theme(legend.position="bottom",legend.text=element_text(size=9))


ggsave('outputs/Equity_Analysis/urban_percent.png')

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=percent_urban_overbudened * 100),color='gray80')+
  scale_fill_distiller(palette='BuPu',direction=1,name='% of Urban Pop.')+
  geom_sf(data=states_2272,fill='transparent',color='gray30')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='black')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()+
  theme(legend.position="bottom",legend.text=element_text(size=9))

ggsave('outputs/Equity_Analysis/overburdened_percent.png')

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=Data_County3,aes(fill=percent_urban_overbudened * 100),color='gray80')+
  geom_sf(data=Data_County3_centroid %>% filter(funding_per_urban_capita > 0),aes(size=funding_per_urban_capita),color='#74c5f7')+
  scale_size_continuous(range = c(0.5,5),name='Funding Per Urban Capita')+
  scale_fill_distiller(palette='Purples',direction=1,name='% of Urban Pop. \n that is Overburdened')+
  geom_sf(data=states_2272,fill='transparent',color='gray30')+
  geom_sf_text(data=Data_County3_centroid %>% filter(NAME != 'New Castle'),aes(label=NAME),size=2,color='black',position=position_nudge(x=40000,y=20000))+
  geom_sf_text(data=Data_County3_centroid %>% filter(NAME == 'New Castle'),aes(label=NAME),size=2,color='black',position=position_nudge(x=40000,y=-20000))+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()

ggsave('outputs/Equity_Analysis/overburdened_percent_funding.png')
```

``` {r export_results}

Data_county3_org <- Data_County3 %>%
  select(NAME,State,funding,subsidy,pop_Urban,funding_per_urban_capita,subsidy_per_capita,percent_urban_pop,percent_urban_area,percent_urban_overbudened) %>%
  st_drop_geometry()

write.csv(Data_county3_org %>% st_drop_geometry(),'results.csv')
  
```

``` {r}

# Aggregate results and export shapefile for mapping in ArcGIS

location_data <- data_raw %>%
  group_by(location_final,Latitude,Longitude) %>% summarize(funding = sum(equity_analysis_funding)) %>%
  st_as_sf(., coords = c("Longitude","Latitude"),crs='EPSG:4326') %>%
  st_transform('EPSG:2272') %>%
  st_join(.,basin_counties,left=FALSE)

st_write(location_data,"outputs/Equity_Analysis/results_for_mapping.shp")

st_write(pop_data_basin,"outputs//Equity_Analysis/census_block_data.shp")

```

``` {r map}

ggplot()+
  geom_sf(data=states_2272,size=2,fill='gray95')+
  geom_sf(data=basin_counties,fill='white')+
  geom_sf(data=pop_data_basin %>% filter(overburdened == 'Yes'),aes(fill='#66c2a5'),color='transparent')+
  geom_sf(data=pop_data_basin %>% filter(overburdened == 'Yes' & urban == 'Urban'),aes(fill='#fff39b'),color='transparent')+
  geom_sf(data=pop_data_basin %>% filter(overburdened == 'No' & urban == 'Urban'),aes(fill='#fc8d62'),color='transparent')+
  geom_sf(data=basin_counties,color='gray80',fill='transparent')+
  geom_sf(data=states_2272,fill='transparent',color='gray60')+
  geom_sf_text(data=basin_counties,aes(label=NAME),size=2,color='black')+
  scale_fill_manual(values=c('#66c2a5','#fff39b','#fc8d62'),label=c('CEJST Rural','Non CEJST Urban','CEJST Urban'),name='')+
  xlim(xmin,xmax)+
  ylim(ymin,ymax)+
  theme_void()

ggsave('outputs/Equity_Analysis/overburdened_communities.png')

```
